---
title: "VTPEH 6270 - Data Simulation II"
author: "Amandine Gamble"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '1'
  html_document:
    toc: true
    toc_depth: 1
    number_sections: true
editor_options:
  chunk_output_type: console
subtitle: ''
urlcolor: blue
---

```{r markdown-options, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

```{r setup, results = 'hide'}

library(ggplot2)
library(dplyr)
library(tidyr)

```

# Simulation 

We simulate the relationship between cheese consumption (recorded as low or high) and happiness (recorded as score). We use the following parameters:

* $n$ (nn): sample size
* $p_c$ (prob_high_cheese): probability of high cheese consumption
* $h_{c}$ (score_happy_given_high_cheese): mean happiness score given high cheese consumption
* $h_{\overline{c}}$ (score_happy_given_low_cheese): mean happiness score given low cheese consumption
* $h_{sd}$ (happy_sd): standard deviation of happiness score within each group

## Function

```{r}

f_sim_cheese = function(nn, 
                        prob_high_cheese,
                        score_happy_given_high_cheese, 
                        score_happy_given_low_cheese, 
                        happy_sd
                        ){
  
  # Cheese vector
  v_cheese = sample(
    c("High", "Low"),
    nn,
    replace = TRUE,
    prob = c(prob_high_cheese, 1 - prob_high_cheese)
  )
  
  # Happiness vector
  v_happiness = NA
  for (jj in 1:nn) {
    if (v_cheese[jj] == "High") {
      v_happiness[jj] = rnorm(1, score_happy_given_high_cheese, happy_sd)
    } else if (v_cheese[jj] == "Low") {
      v_happiness[jj] = rnorm(1, score_happy_given_low_cheese, happy_sd)
    }
  }

  # Combine data
  data = data.frame(
    cheese = v_cheese, 
    happiness = v_happiness)
  
  return(data)
}

```

## Runs

```{r}

# Parameters for tests
# prob_high_cheese = 0.4
# score_happy_given_high_cheese = 0.8
# score_happy_given_low_cheese = 0.3
# happy_sd = 0.01

# Initialize objects
list_datasets <- list()
df_stats <- data.frame(beta = NA, 
                       sample = NA, 
                       sim = NA, 
                       high_cheese_mean = NA, 
                       high_cheese_sd = NA, 
                       low_cheese_mean = NA, 
                       low_cheese_sd = NA,
                       p = NA)
sim_index = 1
score_happy_given_high_cheese = 0.8

# Loop on ???
for (beta in seq(0, 0.2, 0.01)) {
  
  score_happy_given_low_cheese = score_happy_given_high_cheese - beta
  
  # Loop on ???
  for (nn in seq(40, 100, 10)) {
    
    # Run simulation 100 times
    for (ii in 1:100) {
      
      # Generate data
      sim_data <- f_sim_cheese(
        nn = nn,
        prob_high_cheese = 0.4,
        score_happy_given_high_cheese = score_happy_given_high_cheese,
        score_happy_given_low_cheese = score_happy_given_low_cheese,
        happy_sd = 0.2
      )
      
      # Store full dataset (comment to go faster)
      #list_datasets[[sim_index]] <- sim_data
      
      # Calculate mean and sd for each cheese group
      df_stats[sim_index, ] = NA
      df_stats[sim_index, "beta"] = beta
      df_stats[sim_index, "sample"] = nn
      df_stats[sim_index, "sim"] = ii
      df_stats$high_cheese_mean[sim_index] =
        round(mean(sim_data[sim_data$cheese == "High", "happiness"]), 2)
      df_stats$high_cheese_sd[sim_index] =
        round(sd(sim_data[sim_data$cheese == "High", "happiness"]), 2)
      df_stats$low_cheese_mean[sim_index] =
        round(mean(sim_data[sim_data$cheese == "Low", "happiness"]), 2)
      df_stats$low_cheese_sd[sim_index] =
        round(sd(sim_data[sim_data$cheese == "Low", "happiness"]), 2)
      
      # T-test
      test = t.test(happiness ~ cheese, sim_data)
      df_stats$p[sim_index] <- test$p.value
      
      # ???
      sim_index = sim_index + 1
      
    }
    
  }
  
}

# Clean environment 
remove(ii, nn, beta, sim_index, test, sim_data)

# Check: 1 simulation run
#head(list_datasets[[1]])

# Check: compiled summary statistics
head(df_stats)

# Save simulation summary
write.csv(df_stats, "Export/sim_summary.csv", row.names = F, quote = F)

# Pivot stats
df_stats_long <- df_stats %>%
  select(beta, sample, sim, high_cheese_mean, low_cheese_mean) %>%
  pivot_longer(cols = c(high_cheese_mean, low_cheese_mean),
               names_to = "group",
               values_to = "mean_value") %>%
  mutate(group = ifelse(group == "high_cheese_mean", "High Cheese", "Low Cheese"))

```

# Exploration

## Summary statistics

```{r}

p1 = ggplot(df_stats_long, aes(x = mean_value, fill = group)) +
  geom_histogram(alpha = 0.7,
                 bins = 30,
                 position = "identity") +
  facet_grid(group ~ .) +
  labs(x = "Mean Happiness Score", y = "Frequency", fill = "Cheese Group") +
  theme_minimal() +
  theme(legend.position = "none") 

p1

```

## P-value

```{r}

p2 = ggplot(df_stats, aes(x = p)) +
  geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
  labs(x = "p-value",
       y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none") 

p2

```

```{r}

p3 = ggplot(df_stats, aes(x = beta, y = p, color = sample)) +
  geom_jitter() +
  labs(x = "Effect size",
       y = "p-value") +
  scale_color_gradient(name = "Sample size") +
  theme_minimal() 

p3

```

## Power

```{r}

power_results <- df_stats %>%
  group_by(beta, sample) %>%
  summarise(
    power = sum(p < 0.05, na.rm = TRUE)/sum(!is.na(p)),
    .groups = "drop"
  )

```

# Visualization

First try...

```{r}

ggplot(power_results, aes(x = beta, y = sample, fill = power)) +
  geom_tile() +
  labs(x = "Effect size",
       y = "Sample size") 

```

Final version...

```{r}

p4 = ggplot(power_results, aes(x = beta, y = sample, fill = power)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Power", 
                       labels = scales::percent_format()) +
  geom_contour(aes(z = power), breaks = 0.8, color = "white", size = 1) +
  labs(x = "Effect size",
       y = "Sample size",
       title = "Statistical Power Analysis",
       subtitle = "White contour line indicates 80% power") +  # Fixed subtitle
  theme_minimal()

p4

```

Composed plot

```{r}

library(patchwork)

combined_plot <- ((p1 / p2) | (p3 / p4)) + 
  plot_annotation(
    title = "Simulation Study Results: Cheese Consumption and Happiness",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5))
  ) 

combined_plot

```

